---
title: "STAT 420 Project"
author: "Group 9"
date: "11/10/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Introduction to the data set and research problems

The COVID-19 pandemic has dramatically affected the daily lives of everyone in the world. As such, our group wanted to investigate the toll the pandemic has taken on different places across the globe. The dataset we have selected consists of worldwide COVID-19 data, broken down for each country, and broken down for each state within the US. The attributes within this data set are confirmed cases, deaths, recoveries, the rate of increase, and their corresponding dates. Within this dataset, we plan to use both the countries-aggregated.csv and worldwide-aggregated.csv to combine all the attributes and discover the relationships and trends within this combined data set.

Specifically, we want to research a few areas like, are confirmed COVID-19 cases correlated with deaths? And if so, is there a lag between high confirmed COVID-19 cases and high death cases? Additionally, we want to look at how the trend of COVID-19 cases change based on geographic location around the world. This delves into how strict countries were in quarantining and lockdown, so it would be interesting to see if there was an optimal quarantining method. Furthermore, we also want to look at if we can predict future COVID-19 cases using forecasting. Finally, we want to look at the times of the year where COVID-19 cases and deaths occur the most and if it is seasonal.

##Exploratory Data Analysis

```{r reading in data}
library(readr)
library(ggplot2)
library(astsa)
library(lubridate)
library(tidyverse)

covid = read_csv("data/worldwide-aggregate.csv")
countries<- read_csv("data/countries-aggregated.csv")
world <- read_csv("data/worldwide-aggregate.csv")
```

```{r exploratory}
ggplot(covid, aes(x=Date, y=Deaths)) +
  geom_line() + xlab("Date") + ylab("Deaths") + 
  ggtitle("COVID Deaths Over Time")

ggplot(covid, aes(x=Date, y=Confirmed)) +
  geom_line() + xlab("Date") + ylab("Confirmed Cases") + 
  ggtitle("COVID Confirmed Cases Over Time")
#We can see that over time the deaths are increasing as well as the confirmed COVID cases. 
```


```{r detrending through regression and differencing}
y=world$Confirmed
x=world$Deaths
time=world$Date

y.lm=lm(y~time)
plot(world$Date,y, type='l', xlab='time', ylab='Confirmed')
abline(reg=y.lm)

x.lm=lm(x~time)
plot(world$Date,x, type='l', xlab='time', ylab='Deaths')
abline(reg=x.lm)

par(mfrow=c(2,2))
res=resid(x.lm)
plot(res, type='l', ylab = 'residual', main='Residual plot')
plot(diff(res), type='l', ylab="difference", main = 'First order differencing')
acf(res, main='Residual')
acf(diff(res), main="Differenced residual")

par(mfrow=c(2,2))
res=resid(y.lm)
plot(res, type='l', ylab = 'residual', main='Residual plot')
plot(diff(res), type='l', ylab="difference", main = 'First order differencing')
acf(res, main='Residual')
acf(diff(res), main="Differenced residual")


```

```{r first, second, and seasonal differencing }
#1st difference, acf, pacf
par(mfrow=c(2,2))
acf(diff(y), main='ACF Confirmed')
pacf(diff(y), main='PACF Confirmed')
acf(diff(x), main='ACF Deaths')
pacf(diff(x), main='PACF Deaths')

par(mfrow=c(1,3))
#plot 2nd difference, acf, pacf
plot(diff(diff(y)), type='l')
acf(diff(diff(y)))
pacf(diff(diff(y)))

#plot 1st seasonal difference, acf, pacf
plot(diff(diff(diff(y)), 7))
acf(diff(diff(diff(y)), 7))
fin = diff(diff(diff(y)), 7)
pacf(fin)

```

```{r moving averages}
# Moving averages - day 50, 100, and 200:
covid$Date <- as.Date(covid$Date, "%Y-%m-%d")
plot(covid$Date, covid$Deaths, xaxt="n", type="l", col="gray", lwd=2)
axis(1, covid$Date, format(covid$Date, "%m-%Y"))
lines(covid$Date, zoo::rollmean(covid$Deaths, 50, align='right', fill=NA), col='red', lwd=2)
lines(covid$Date, zoo::rollmean(covid$Deaths, 100, align='right', fill=NA), col='blue', lwd=2)
lines(covid$Date, zoo::rollmean(covid$Deaths, 200, align='right', fill=NA), col='green', lwd=2)

```

```{r exploratory model analysis}
library(xts)
covid_ts = xts(covid$Deaths, as.Date(covid$Date, format='%Y-%m-%d'))

par(mfrow=c(2,2))
plot(covid_ts)
plot(log(covid_ts))
plot(diff(log(covid_ts)))
plot(diff(diff(log(covid_ts))))
y = diff(diff(log(covid_ts)))

par(mfrow=c(1, 2))
acf(y, lag.max=10, na.action=na.pass)
pacf(y, lag.max=10, na.action=na.pass)


```
From this we can see that the data is not normally distributed and it becomes stationary after being differenced twice. We can also see that there is no clear cut off for the ACF nor PACF. 

```{r MA model}
MA = arima(covid_ts, order = c(0, 0, 1))
MA_fit = covid_ts - residuals(MA)
plot.ts(covid_ts, ylab="Deaths", main="COVID Deaths Over Time", lwd=2)
points(MA_fit, type = "l", col = "red", lty = 2, lwd=2)

MA_diff = arima(covid_ts, order = c(0, 1, 1))
MA_diff_fit = covid_ts - residuals(MA_diff)
plot.ts(covid_ts, ylab="Deaths", main="COVID Deaths Over Time", lwd=2)
points(MA_diff_fit, type = "l", col = "green", lty = 2, lwd=2)
```
For the MA model we look at the date vs deaths. We can see that the MA(1) fit is not very good and that it looks much better after differencing. 

```{r AR model}
plot.ts(covid_ts, ylab="Deaths", main="COVID Deaths Over Time", lwd=2)
AR <- arima(covid_ts, order = c(1,0,0))
AR_fit <- covid_ts - residuals(AR)
points(AR_fit, type = "l", col = "red", lty = 2, lwd=2)

predict_AR <- predict(AR, n.ahead=12)
predict_AR$pred



```
For the AR model we also look at the date vs deaths. We can see that the AR(1) fit looks pretty good but compare it to the MA model.

```{r MA vs AR}

AIC(AR) # 13,103.2
AIC(MA_diff)  # 12,398.28
AIC(MA) # 18,712.88

par(mfrow=c(2,2))
plot(MA_diff$residuals)
plot(AR$residuals)

hist(MA_diff$residuals)
qqnorm(MA_diff$residuals)
qqline(MA_diff$residuals)
```
We can also see that the residuals have some trend but look fairly normal. By comparing the AICs we can see that the MA(1) with differencing = 1 has the lowest AIC. However, the AICs are fairly large leading us to believe that neither MA or AR models are fit for this data. 

##Statistical results and conclusions from models and methods 

# Research Question #1: Are confirmed cases linearly correlated with deaths?

```{r RQ1}
cor(covid$Confirmed, covid$Deaths)
plot(covid$Confirmed, covid$Deaths, xlab="Confirmed Cases", ylab="Deaths", main="COVID Confirmed Cases vs Deaths", type='l')

```
Yes, confirmed cases and deaths appear to have a linear relationship and have a high correlation of 0.997. 

# Research Question #2: How does the trend of covid cases change based on geographic location?

```{r RQ2}

### North American Countries
us_country <- countries[countries$Country == "US",]
canada <- countries[countries$Country == "Canada",]
mexico <- countries[countries$Country == "Mexico",]

us_country$Date=with_tz(us_country$Date, "America/New_York")
canada$Date=with_tz(canada$Date, "America/New_York")
mexico$Date=with_tz(mexico$Date, "America/New_York")

us_country$per = us_country$Confirmed/100000
canada$per = canada$Confirmed/100000
mexico$per = mexico$Confirmed/100000

(ggplot(us_country, aes(x = Date, y = `per`)) +
  geom_line(aes(y = `per`, color = "USA"))  +
  geom_line(data = canada, aes(color = "Canada")) +
  geom_line(data = mexico, aes(color = "Mexico")) +
  xlab('Date') +
  ylab('Count') +
  ggtitle('North American Country Covid Cases per 100,000'))

### Asian Countries
india <- countries[countries$Country == "India",]
china <- countries[countries$Country == "China",]
japan <- countries[countries$Country == "Japan",]

india$Date=with_tz(india$Date, "America/New_York")
china$Date=with_tz(china$Date, "America/New_York")
japan$Date=with_tz(japan$Date, "America/New_York")

india$per = india$Confirmed/100000
china$per = china$Confirmed/100000
japan$per = japan$Confirmed/100000


(ggplot(india, aes(x = Date, y = `per`)) +
  geom_line(aes(y = `per`, color = "India"))  +
  geom_line(data = china, aes(color = "China")) +
  geom_line(data = japan, aes(color = "Japan")) +
  xlab('Date') +
  ylab('Count') +
  ggtitle('Asian Country Covid Cases per 100,000'))

```

As seen by the graphs, there seems to be a high count of COVID-19 cases in the United States and India. While these two countries are in different geographic locations, it seems as though the quarantining policy wasn't strict enough, wasn't enforced properly, or wasn't followed in these two countries. Further research will need to be done to see if there is a change in trend based on geographic location. However, it is more likely based on this analysis that a refinement to the research question could be made to measure the difference in COVID-19 case trends in countries based on the strictness of their quarantining policies.


# Research Question #3: How can we predict the number of covid cases with forecasting?
```{r RQ3 model 1}
confirmed.arima1 = arima(y, order=c(0, 2,1), seasonal=list(order=c(0, 1,1), period=7))

plot(confirmed.arima1$residuals, ylab = 'residuals')
acf(confirmed.arima1$residuals, lag.max  = 50, na.action=na.pass)
pacf(confirmed.arima1$residuals, lag.max  = 50, na.action=na.pass)

qqnorm(confirmed.arima1$residual)
qqline(confirmed.arima1$residual)
Box.test(confirmed.arima1$residuals, fitdf=1, lag = 20, type="Ljung")

confirmed.arima1$aic
```

```{r RQ3 model 2}
confirmed.arima2 = arima(y, order=c(0, 2,2), seasonal=list(order=c(0, 1,1), period=7))

plot(confirmed.arima2$residuals, ylab = 'residuals')
acf(confirmed.arima2$residuals, lag.max  = 50, na.action=na.pass)

qqnorm(confirmed.arima2$residual)
qqline(confirmed.arima2$residual)

confirmed.arima2$aic
Box.test(confirmed.arima2$residuals, lag = 20, fitdf=2, type="Ljung")

```
We found that model 1 would be the best which is the ARIMA(0,2,1)x(0,1,1)7. Next, we take a closer look at the overall forecasting using the model. 

```{r RQ3 forecast}

#Forecasting with model 1 confirmed cases
forecast=predict(confirmed.arima1, n.ahead =50)

#overall forecast
plot(y, xlim = c(0,700), ylim = c(0, 262659973), type='l')
lines(forecast$pred, col="green")

#closer look
sarima.for(world$Confirmed, 50, 1,2,0,1,1,0,7)

```

```{r RQ3 exploratory}
#initial plot
ts.plot(world$Recovered[0:500])
x = world$Recovered[0:500]
par(mfrow=c(1,2))
acf(x)
pacf(x)

#1st difference plot, acf, pacf
par(mfrow=c(1, 1))
diff(x)
ts.plot(diff(x))
par(mfrow=c(1, 2))
acf(diff(x))
pacf(diff(x))

#2st difference plot, acf, pacf
par(mfrow=c(1, 1))
ts.plot(diff(diff(x)))
par(mfrow=c(1, 2))
acf(diff(diff(x)))
pacf(diff(diff(x)))

```

```{r RQ3 arima3 & arima4}
par(mfrow=c(2, 2))
confirmed.arima3 = arima(x, order=c(0, 2, 1))
plot(confirmed.arima3$residuals, ylab = 'residuals')
acf(confirmed.arima3$residuals, lag.max  = 50)
pacf(confirmed.arima3$residuals, lag.max  = 50)

qqnorm(confirmed.arima3$residual)
qqline(confirmed.arima3$residual)

Box.test(confirmed.arima3$residuals, lag = 20, fitdf=1, type="Ljung")

confirmed.arima3$aic

confirmed.arima4 = arima(x, order=c(0, 2, 2))
plot(confirmed.arima4$residuals, ylab = 'residuals')
acf(confirmed.arima4$residuals, lag.max  = 50)
pacf(confirmed.arima4$residuals, lag.max  = 50)

qqnorm(confirmed.arima4$residual)
qqline(confirmed.arima4$residual)

Box.test(confirmed.arima4$residuals, lag = 20, fitdf=1, type="Ljung")

confirmed.arima4$aic

```
We can see that the ARIMA(0,2,1) model is the best, so we forecast with this model

```{r RQ3 forecast2}
##Forecasting with recovered cases
forecast=predict(confirmed.arima3, n.ahead =50)

plot(x, type = 'l', xlim = c(0,550), ylim=c(0,139575581))
lines(forecast$pred, col ='green')

#closer look
sarima.for(x,30, 0,2,1, 0,0,0,0)

```

```{r RQ3 exploratory continued}
#de-trend deaths
ts.plot(world$Deaths)
d = world$Deaths
par(mfrow=c(1, 2))
acf(d)
pacf(d)

#1st difference plot, acf, pacf
par(mfrow=c(1, 1))
ts.plot((diff(d)))
par(mfrow=c(1, 2))
acf(diff(d), lag.max=(50))
pacf(diff(d),  lag.max=(50))

#2nd difference plot, acf, pacf
par(mfrow=c(1, 1))
ts.plot((diff(diff(d))))
par(mfrow=c(1, 2))
acf(diff(diff(d)), lag.max=(50))
pacf(diff(diff(d)),  lag.max=(50))

#1st seasonal difference period 7 plot, acf, pacf
acf(diff(diff(diff(d)), 7), lag.max=50)
pacf(diff(diff(diff(d)), 7), lag.max=50)
```


```{r RQ3 arima5 & arima6}
confirmed.arima5 = arima(d, order=c(0, 1, 1),  seasonal=list(order=c(0, 1,2), period=7))
par(mfrow=c(1, 1))
plot(confirmed.arima5$residuals, ylab = 'residuals')
par(mfrow=c(1, 2))
acf(confirmed.arima5$residuals, lag.max  = 50)
pacf(confirmed.arima5$residuals, lag.max  = 50)

qqnorm(confirmed.arima5$residual)
qqline(confirmed.arima5$residual)

Box.test(confirmed.arima5$residuals, lag = 20,fitdf=1)
confirmed.arima5$aic

confirmed.arima6 = arima(d, order=c(0, 1, 1),  seasonal=list(order=c(0, 1,1), period=7))
par(mfrow=c(1, 1))
plot(confirmed.arima6$residuals, ylab = 'residuals')
par(mfrow=c(1, 2))
acf(confirmed.arima6$residuals, lag.max  = 50)
pacf(confirmed.arima6$residuals, lag.max  = 50)

qqnorm(confirmed.arima6$residual)
qqline(confirmed.arima6$residual)

Box.test(confirmed.arima6$residuals, lag = 20,fitdf=1)
confirmed.arima6$aic


```
We can see that the ARIMA(0,1,1)x(0,1,2)7 model is the best so we use it to forecast deaths

```{r RQ3 forecast3}
forecast=predict(confirmed.arima5, n.ahead =50)

plot(d, type = 'l', ylim= c(0, 5236472), xlim= c(0,700))
lines(forecast$pred, col ='green')

#closer look
sarima.for(d, 50, 0,1,1,0,1,2,7)
```

The overall conclusion regarding this question is the confirmed cases and deaths both can be forecasted using an ARIMA(0,2,1)x(0,1,1)7 model. However, recoveries doesn't seem to have a seasonal component so it can be modeled with ARIMA(0,2,1). The period of 7 for seasonality of confirmed cases and deaths can be explained by the trends we see weekly and the way the data is collected. Oftentimes there are patterns depending on the day of the week. 

# Research Question #4: What times of the year do the most covid cases and/or deaths occur?

#Worldwide Confirmed Cases
worldwide_data <- read.csv("~/stat420/covid_data/worldwide-aggregate.csv")
worldwide_data$Date=as.Date(with_tz(worldwide_data$Date, "America/New_York"))
ggplot(worldwide_data, aes(x=Date, y=Confirmed))+
  geom_line(color = 'orange')+
  labs(x='time', y='# Deaths', title='Worldwide Confirmed Cases')+
  scale_x_date(date_breaks = "2 months" , date_labels = "%b-%y")

a = (diff(diff(worldwide_data$Confirmed)))
plot.ts(a)
acf(a)
pacf(a)
ww_arima = arima(worldwide_data$Confirmed, order= c(0,2,0))

#Worldwide Increase Rate
worldwide_data$Date=as.Date(with_tz(worldwide_data$Date, "America/New_York"))
ggplot(worldwide_data, aes(x=Date, y=Increase.rate))+
  geom_line(color = 'orange')+
  labs(x='time', y='# Deaths', title='Worldwide Confirmed Cases')+
  scale_x_date(date_breaks = "2 months" , date_labels = "%b-%y")


Based on the plots, we can see there is a slight seasonal trend with Confirmed COVID cases on the worldwide scale approximately every 3 months. There is the initial spike of increase rate at the start of the pandemic and another notable drastic increase in December 2020. 


